---
output: 
  html_document: 
    toc: yes
    fig_caption: yes
    toc_float: true
  pdf_document: 
    toc: yes
    fig_caption: yes
title: "Movie Recommendation System using ML in R"
author: Vatsal Desai
abstract: "The main goal of this machine learning project is to build a recommendation engine that recommends movies to users. We will be developing an Item Based Collaborative Filter using recommenderlab package. "
date: "`r format(Sys.time(), '%B %d %Y')`"
---
```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.align = 'center'
)
```

\newpage  
  
# Introduction  
  
## What is a Recommendation System  
  
A recommendation system provides suggestions to the users through a filtering process that is based on user preferences and browsing history. The information about the user is taken as an input. The information is taken from the input that is in the form of browsing data. This information reflects the prior usage of the product as well as the assigned ratings. A recommendation system is a platform that provides its users with various contents based on their preferences and likings. A recommendation system takes the information about the user as an input. The recommendation system is an implementation of the machine learning algorithms.  
A recommendation system also finds a similarity between the different products. For example, Netflix Recommendation System provides you with the recommendations of the movies that are similar to the ones that have been watched in the past. Furthermore, there is a collaborative content filtering that provides you with the recommendations in respect with the other users who might have a similar viewing history or preferences. There are two types of recommendation systems – Content-Based Recommendation System and Collaborative Filtering Recommendation. In this project of recommendation system in R, we will work on a collaborative filtering recommendation system and more specifically, ITEM based collaborative recommendation system.  
  
# Dataset and libraries  
  
In order to build our recommendation system, we have used the *MovieLens* Dataset. You can find the *movies.csv* and *ratings.csv* file that we have used in our Recommendation System Project [**here**](https://drive.google.com/file/d/1Dn1BZD3YxgBQJSIjbfNnmCFlDW2jdQGD/view) This data consists of 105339 ratings applied over 10329 movies.  
```{r}
library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)
```
```{r}
movie_data <- read.csv("movies.csv", stringsAsFactors = FALSE)
rating_data <- read.csv("ratings.csv")
str(movie_data)
```
We can overview the summary of the movies using the *summary()* function. We will also use the *head()* function to print the first six lines of *movie_data*  
```{r}
summary(movie_data)
head(movie_data)
```
Similarly, we can output the summary as well as the first six lines of the ‘rating_data’ dataframe
```{r}
summary(rating_data)
head(rating_data)
```
  
# Data pre-processing  
  
From the above table, we observe that the *userId* column, as well as the *movieId* column, consist of integers. Furthermore, we need to convert the genres present in the *movie_data* dataframe into a more usable format by the users. In order to do so, we will first create a *one-naught encoding* to create a matrix that comprises of corresponding genres for each of the films.  
```{r}
movie_genre <- as.data.frame(movie_data$genres,stringsAsFactors = FALSE)
movie_genre2 <- as.data.frame(tstrsplit(movie_genre[,1], '[|]', type.convert = TRUE), 
                              stringAsFactors = FALSE)
colnames(movie_genre2) <- c(1:10)

list_genre <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")
genre_mat1 <- matrix(0,10330,18)
genre_mat1[1,] <- list_genre
colnames(genre_mat1) <- list_genre

for (index in 1:nrow(movie_genre2)) {
  for (col in 1:ncol(movie_genre2)) {
    gen_col = which(genre_mat1[1,] == movie_genre2[index,col])
    genre_mat1[index+1,gen_col] <- 1
}
}

genre_mat2 <- as.data.frame(genre_mat1[-1,], stringsAsFactors=FALSE)
for (col in 1:ncol(genre_mat2)) {
  genre_mat2[,col] <- as.integer(genre_mat2[,col])
} 
str(genre_mat2)
```
In the next step of Data Pre-processing of data, we will create a *‘search matrix’* that will allow us to perform an easy search of the films by specifying the genre present in our list.  
```{r}
SearchMatrix <- cbind(movie_data[,1:2], genre_mat2[])
head(SearchMatrix)
```
There are movies that have several genres, for example, Toy Story, which is an animated film also falls under the genres of Comedy, Fantasy, and Children. This applies to the majority of the films.  
For our movie recommendation system to make sense of our ratings through *recommenderlabs*, we have to convert our matrix into a sparse matrix one. This new matrix is of the class *‘realRatingMatrix’*. This is performed as follows:  
```{r, cache=TRUE}
ratingMatrix <- dcast(rating_data, userId~movieId, value.var = "rating", na.rm=FALSE)
ratingMatrix <- as.matrix(ratingMatrix[,-1])
ratingMatrix <- as(ratingMatrix, "realRatingMatrix")
ratingMatrix
```
  
# Choosing recommendation model  
  
Let us now overview some of the important parameters that provide us various options for building recommendation systems for movies  
```{r}
recommendation_model <- recommenderRegistry$get_entries(dataType = "realRatingMatrix")
names(recommendation_model)
```
```{r}
lapply(recommendation_model, "[[", "description")
```
Out of all these available models, we shall apply *Item based Collaborative Filtering* model to our data here.  
```{r}
recommendation_model$IBCF_realRatingMatrix$parameters
```
  
# Exploring similar data  
  
Collaborative Filtering involves suggesting movies to the users that are based on collecting preferences from many other users. For example, if a user A likes to watch action films and so does user B, then the movies that the user B will watch in the future will be recommended to A and vice-versa. Therefore, recommending movies is dependent on creating a relationship of similarity between the two users. With the help of *recommenderlab*, we can compute similarities using various operators like cosine, pearson as well as jaccard.  
```{r}
similarity_mat <- similarity(ratingMatrix[1:4, ],
                               method = "cosine",
                               which = "users")
as.matrix(similarity_mat)

image(as.matrix(similarity_mat), main = "User's Similarities")
```
In the above matrix, each row and column represents a user. We have taken four users and each cell in this matrix represents the similarity that is shared between the two users.  
Now, we delineate the similarity that is shared between the films  
```{r}
movie_similarity <- similarity(ratingMatrix[, 1:4], method =
                                 "cosine", which = "items")
as.matrix(movie_similarity)

image(as.matrix(movie_similarity), main = "Movies similarity")
```
Let us now extract the most unique ratings  
```{r}
rating_values <- as.vector(ratingMatrix@data)
unique(rating_values)
```
Now, we will create a table of ratings that will display the most unique ratings.  
```{r, cache=TRUE}
Table_of_Ratings <- table(rating_values)
Table_of_Ratings
```
  
# Most viewed movies visualization  
  
In this section of the machine learning project, we will explore the most viewed movies in our dataset. We will first count the number of views in a film and then organize them in a table that would group them in descending order.  
```{r, cache=TRUE}
movie_views <- colCounts(ratingMatrix)
table_views <- data.frame(movie = names(movie_views), views = movie_views)
table_views <- table_views[order(table_views$views, decreasing = TRUE), ]
table_views$title <- NA
for (index in 1:10325){
  table_views[index,3] <- as.character(subset(movie_data,
                                         movie_data$movieId == table_views[index,1])$title)
}
table_views[1:6,]
```
Now, we will visualize a bar plot for the total number of views of the top films. We will carry this out using *ggplot2*  
```{r, cache=TRUE}
ggplot(table_views[1:6, ], aes(x = title, y = views)) +
  geom_bar(stat="identity", fill = 'steelblue') +
  geom_text(aes(label=views), vjust=-0.3, size=3.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +

  ggtitle("Total Views of the Top Films")
```
From the above bar-plot, we observe that Pulp Fiction is the most-watched film followed by Forrest Gump.  
  
# Heatmap of movie ratings  
  
Now, we shall visualize a heatmap of the movie ratings. This heatmap will contain first 25 rows.  
```{r}
image(ratingMatrix[1:20, 1:25], axes = FALSE, main = "Heatmap of the first 25 rows and 25 columns")
```
  
# Data preparation  
  
We shall conduct the data preparation in the following 3 steps.  
- Selecting useful data  
- Normalizing data  
- Binarizing the data  
  
## Finding useful data  
  
For finding useful data in our dataset, we have set the threshold for the minimum number of users who have rated a film as 50. This is also same for minimum number of views that are per film. This way, we have filtered a list of watched films from least-watched ones.  
```{r}
movie_ratings <- ratingMatrix[rowCounts(ratingMatrix) > 50,
                             colCounts(ratingMatrix) > 50]
movie_ratings
```
From the above output of ‘movie_ratings’, we observe that there are 420 users and 447 films as opposed to the previous 668 users and 10325 films. We can now delineate our matrix of relevant users as follows:  
```{r}
minimum_movies<- quantile(rowCounts(movie_ratings), 0.98)
minimum_users <- quantile(colCounts(movie_ratings), 0.98)
image(movie_ratings[rowCounts(movie_ratings) > minimum_movies,
                     colCounts(movie_ratings) > minimum_users],
main = "Heatmap of the top users and movies")
```
Now, we will visualize the distribution of the average ratings per user.  
```{r}
average_ratings <- rowMeans(movie_ratings)
qplot(average_ratings, fill=I("steelblue"), col=I("red")) +
  ggtitle("Distribution of the average rating per user")
```
  
## Data normalization  
  
In the case of some users, there can be high ratings or low ratings provided to all of the watched films. This will act as a bias while implementing our model. In order to remove this, we normalize our data.  
Normalization is a data preparation procedure to standardize the numerical values in a column to a common scale value. This is done in such a way that there is no distortion in the range of values. Normalization transforms the average value of our ratings column to 0. We then plot a heatmap that delineates our normalized ratings.  
```{r}
normalized_ratings <- normalize(movie_ratings)
sum(rowMeans(normalized_ratings) > 0.00001)

image(normalized_ratings[rowCounts(normalized_ratings) > minimum_movies,
                          colCounts(normalized_ratings) > minimum_users],
main = "Normalized Ratings of the Top Users")
```
  
## Data binarization  
  
In the final step of our data preparation, we will binarize our data. Binarizing the data means that we have two discrete values 1 and 0, which will allow our recommendation systems to work more efficiently. We will define a matrix that will consist of 1 if the rating is above 3 and otherwise it will be 0.
```{r}
binary_minimum_movies <- quantile(rowCounts(movie_ratings), 0.95)
binary_minimum_users <- quantile(colCounts(movie_ratings), 0.95)
good_rated_films <- binarize(movie_ratings, minRating = 3)
image(good_rated_films[rowCounts(movie_ratings) > binary_minimum_movies,
colCounts(movie_ratings) > binary_minimum_users],
main = "Heatmap of the top users and movies")
```
  
# Collaborative filtering system  
  
Next we shall develop *Item Based Collaborative Filtering* system. This type of collaborative filtering finds similarity in the items based on the people’s ratings of them. The algorithm first builds a similar-items table of the customers who have purchased them into a combination of similar items. This is then fed into the recommendation system.  
We will build this filtering system by splitting the dataset into 80% training set and 20% test set.  
```{r}
sampled_data<- sample(x = c(TRUE, FALSE),
                      size = nrow(movie_ratings),
                      replace = TRUE,
                      prob = c(0.8, 0.2))
training_data <- movie_ratings[sampled_data, ]
testing_data <- movie_ratings[!sampled_data, ]
```
  
# Building recommendation system  
  
We will now explore the various parameters of our Item Based Collaborative Filter. These parameters are default in nature. In the first step, k denotes the number of items for computing their similarities. Here, k is equal to 30. Therefore, the algorithm will now identify the k most similar items and store their number. We use the cosine method which is the default one but, one can also use pearson method.  
```{r}
recommendation_system <- recommenderRegistry$get_entries(dataType ="realRatingMatrix")
recommendation_system$IBCF_realRatingMatrix$parameters
```
```{r}
recommen_model <- Recommender(data = training_data,
                          method = "IBCF",
                          parameter = list(k = 30))
recommen_model
class(recommen_model)
```
Let us now explore our recommendation system model as follows:  
Using the *getModel()* function, we will retrieve the *recommen_model*. We will then find the class and dimensions of our similarity matrix that is contained within *model_info*. Finally, we will generate a heatmap, that will contain the top 20 items and visualize the similarity shared between them.  
```{r}
model_info <- getModel(recommen_model)
class(model_info$sim)
dim(model_info$sim)
top_items <- 20
image(model_info$sim[1:top_items, 1:top_items],
   main = "Heatmap of the first rows and columns")
```
Next, we shall carry out the sum of rows and columns with the similarity of the objects above 0. We will visualize the sum of columns through a distribution as follows:  
```{r}
sum_rows <- rowSums(model_info$sim > 0)
table(sum_rows)
sum_cols <- colSums(model_info$sim > 0)
qplot(sum_cols, fill=I("steelblue"), col=I("red"))+ ggtitle("Distribution of the column count")
```
We will create a *top_recommendations* variable which will be initialized to 10, specifying the number of films to each user. We will then use the *predict()* function that will identify similar items and will rank them appropriately. Here, each rating is used as a weight. Each weight is multiplied with related similarities. Finally, everything is added in the end.  
```{r}
top_recommendations <- 10
predicted_recommendations <- predict(object = recommen_model,
                          newdata = testing_data,
                          n = top_recommendations)
predicted_recommendations
```
```{r}
user1 <- predicted_recommendations@items[[1]]
movies_user1 <- predicted_recommendations@itemLabels[user1]
movies_user2 <- movies_user1
for (index in 1:10){
  movies_user2[index] <- as.character(subset(movie_data,
                                         movie_data$movieId == movies_user1[index])$title)
}
movies_user2
```
```{r}
recommendation_matrix <- sapply(predicted_recommendations@items,
                      function(x){ as.integer(colnames(movie_ratings)[x]) })
recommendation_matrix[,1:4]
```
This also allows us to see how many items are matching to each recommendation in our model and also we can visualizing which of the movies match most of the things in the sample we are working with and how many items they match with based on rating and genre.  
```{r}
number_of_items <- factor(table(recommendation_matrix))
chart_title <- "Distribution of the Number of Items for IBCF"
qplot(number_of_items, fill=I("steelblue"), col=I("red")) + ggtitle(chart_title)
```
```{r}
number_of_items_sorted <- sort(number_of_items, decreasing = TRUE)
number_of_items_top <- head(number_of_items_sorted, n = 4)
table_top <- data.frame(as.integer(names(number_of_items_top)), number_of_items_top)
for(i in 1:4) {
  table_top[i,1] <- as.character(subset(movie_data,movie_data$movieId == table_top[i,1])$title)
}
colnames(table_top) <- c("Movie Title", "No. of Items")
head(table_top)
```
  
# Summary  
  
Recommendation Systems are the most popular type of machine learning applications; that are used in all sectors. They are an improvement over the traditional classification algorithms as they can take many classes of input and provide similarity ranking based algorithms to provide the user with accurate results. These recommendation systems have evolved over time and have incorporated many advanced machine learning techniques to provide the users with the content that they want.  
  
***  
  
\newpage  
  
To recreate any of the above performed operations here; the list of all packages and environment conditions are given below.  
```{r}
sessionInfo()
```
***  